---
title: "00.StudyAreaPrep"
author: "Angela Moreras"
date: "2024-02-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


In this document will have the study area shapefile, along with the BCR shapefile use in the National Models v5, to avoid using different sources for the BCRs delimitations.

Load packages

```{r}
library(tidyverse) #basic data wrangling
library(terra) #basic raster handling
library(sf) #basic shapefile handling
```

#Set folders
```{r}
root<- "."
```

#BCR

Fist set the right projections
```{r}
projection.trans<-5072
projection.st_as_sf<-4326
```

Read in BCR shapefile and select the BCRs needed. Plot the map to check it is okay.
```{r}

bcr <- read_sf(file.path("G:/Shared drives/BAM_NationalModels5/Regions/BAM_BCR_NationalModel.shp")) %>% 
  dplyr::filter(subUnit %in%  c(80,81,82,12,14)) %>% 
  st_transform(crs=projection.trans)

ggplot(bcr) +
  geom_sf(aes(fill=factor(subUnit)))
```


#Study Area

Read Study Area shape file

```{r}
root<-"."

StudyArea<-read_sf(file.path(root, "Study_area/BCR81214_Can/BCR81214_Can.shp"))

StudyArea.trans<- st_transform(StudyArea,crs=projection.trans)  #same projection
```

Crop the BCR file to the extent of the Study area

```{r}
bcr.StudyArea <- bcr %>% 
  st_intersection(StudyArea.trans) %>% 
  mutate(country="ca")

ggplot(bcr.StudyArea) +
  geom_sf(aes(fill=factor(subUnit)))
```

Fused the BCR units to create shapefile for the model. 
For the EHJV purposes we are running models for the study area as a unit, while the national models were run by BCR. 
 
```{r}

StudyAreaFused<- bcr.StudyArea %>% 
  summarise() %>% # not an obvious geometry merge tool, but quite effective :)
  mutate(area = st_area(.))

ggplot(StudyAreaFused) +
  geom_sf()
```
NOTE: It seems that is possible to accelerate the process of data extraction  in wildtrax (1.1.) by fusing the BCR subunits. However we did not use this fused map for this purpose 

# Save
In this case I kept the shape file with the BCR borders. 
```{r}
save(bcr.StudyArea, StudyAreaFused, file=paste0(root, "/StudyArea.Rdata"))
st_write(bcr.StudyArea, paste0(root, "/Study_area/StudyArea_wrangled.shp")) 
```
