---
title: "1.2. ebirdDataExtraction"
author: "Angela Moreras"
date: "2024-03-07"
output: html_document
---

This document is based on the workflow produce by Elly Knight to produce BAM's National models v5. 
The first part of the code is a exact copy of the version (until point #7) produce for the National models v5 (01.DataWrangling section B.GET EBIRD DATA)

Note: Data from eBird is download manually after filling a form to get access to it (https://ebird.org/data/download/ebd). Also auk package requires to install Cygwin software in your computer  (https://www.cygwin.com/install.html) at the default folder (otherwise is no going to be detected by R). For further instruction in the use of auk package see https://cornelllabofornithology.github.io/auk/articles/auk.html 

#0. Prep

```{r}
library(tidyverse) #basic data wrangling
library(data.table) #for binding lists into dataframes
library(lubridate) #date wrangling
library(auk) #eBird wrangling
library(purrr) #Mapping functions to read in multiple files
library(sf) #read in & handle study area
library(terra) #raster management
```

```{r}
root <- "C:/Users/Angie/OneDrive - Université Laval/Regional_models/Analisis_RM"
root.drive<- "G:/Shared drives/BAM_NationalModels/NationalModels5.0/Data"

load(file.path(root, "StudyArea","StudyArea.Rdata"))
```
#1. Get list of ebd objects to process

Note: Before running this part of the code ensure that folder ebd_raw and ebd_filtered are created. 
In folder ebd_raw are going to be all the folders containing the ebd data. 

```{r}
auk_set_ebd_path(file.path(root, "eBird/ebd_raw"), overwrite=TRUE) #Set path

ebd.files <- grep(list.files(file.path(root, "/eBird/ebd_raw"),recursive = TRUE, pattern="ebd_*"), pattern='sampling', invert=TRUE, value=TRUE) #find ebd files except the sampling ones
  
```


#2. Filter data

We are only using stationary data with a range duration from 0 to 10 min, to resemble  point count data.
Data is save keeping in the folder ebd_filtered.

```{r}
for(i in 1:length(ebd.files)){
  
  file<-tail(unlist(strsplit(ebd.files[i],"/")),n=1)
  
  # Define filters and save filtered data
  # this may take several hours. 
  filtered <- auk_ebd(ebd.files[i]) %>% 
    auk_protocol("Stationary") %>% 
    auk_duration(c(0, 10)) %>% 
    auk_filter(file=file.path(root, "eBird/ebd_filtered", file), overwrite=TRUE, 
               keep = c("group identifier", "sampling_event_identifier", "scientific name", "common_name",
                        "observation_count", "latitude", "longitude", "locality_type", "observation_date",
                        "time_observations_started", "observer_id", "duration_minutes"))
  
}

```


#3. Wrangle ebird data
Note that for eBird data is no count was made, an 'X' is used to indicate presence. Therefore, here observations with "X" are assume to be 1 individual. 
Hotspots are remove from the data
```{r}
ebd.files.done <- list.files(file.path(root, "eBird", "ebd_filtered"), pattern="ebd_|.txt", full.names=TRUE)

#Read all ebd files as one element
#Note this next line takes a long time to run (couple hours)
raw.ebd <- purrr::map(.x=ebd.files.done, .f=~read_ebd(.)) %>%  
  rbindlist()

#get species scientific name and code
tax.wt <- read.csv(file.path(root.drive, "Lookups/", "lu_species.csv")) %>%  
  mutate(scientific_name = paste(species_genus, species_name)) %>% 
  dplyr::select(scientific_name, species_code)

#columns to select
colnms <- c("source", "organization", "project_id", "sensor", "tag_method", "location_id", "buffer", "latitude", "longitude", "year", "date", "observer_id", "duration", "distance", "species_code", "abundance")

#create the 
wrangle.ebd <- raw.ebd %>% 
  dplyr::filter(locality_type!="H") %>% 
  mutate(source = "eBird",
         organization = "eBird",
         project_id= "eBird",
         sensor="PC",
         tag_method="PC",
         singlesp="n",
         buffer=0,
         date = ymd_hms(paste0(observation_date, time_observations_started)),
         year = year(date),
         distance = Inf,
         abundance = as.numeric(ifelse(observation_count=="X", 1, observation_count))) %>% 
  rename(duration = duration_minutes,
         location_id = sampling_event_identifier) %>% 
  left_join(tax.wt) %>% 
  dplyr::select(all_of(colnms)) %>% 
  dplyr::filter(!is.na(date))
```
#4. Crop data to study site 
```{r}
ebd.shape <- wrangle.ebd %>% 
             distinct(location_id, .keep_all = TRUE)%>% #use only unique locations
                    sf::st_as_sf(coords = c("longitude", "latitude"), crs=projection.st_as_sf) %>% 
                                                              #transform data to shape file
                    st_transform(projection.trans) %>%        #Change projection to match
                    st_crop(bcr.sa) 


ebd.shape.re<-st_intersection(ebd.shape, bcr.sa) #get rid of points outside the Study Area
     
unique.loc<- ebd.shape.re$location_id #which location are in the study area
      
use.ebd<- wrangle.ebd %>% filter(location_id %in% unique.loc)
```

# 5. Save
```{r}
save(use.ebd, file=paste0(root, "/eBird/ebd", Sys.Date(), ".Rdata"))
```
  